
---
title: "Reproducible Research: Peer Assessment 1"
author: "Fallatah K. M."
date:    "9/28/2020"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---

```{r setup, include=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Sys.setenv(TZ = "UTC")      # set global TZ for POSIXt class
.Rfliburl <- "https://raw.githubusercontent.com/robertwcw/Rflib/master"
source(file.path(.Rfliburl,"getRflib.R"),local = TRUE)
source(getRflib("is.defined.R"),local = TRUE)
source(getRflib("myplclust.R"),local = TRUE)
source(getRflib("strCap.R"),local = TRUE)
library(dplyr, warn.conflicts = TRUE, quietly = TRUE)
library(lubridate, warn.conflicts = TRUE, quietly = TRUE)
library(scales, warn.conflicts = TRUE, quietly = TRUE)
```
&nbsp;


```{r, message = FALSE}
filetmp <- tempfile()
datadir <- paste(".", "data", sep = "/")
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileurl, filetmp)
if (!dir.exists(datadir)) 
    {
        dir.create(datadir)
    }
unzip(filetmp, exdir = datadir, overwrite = TRUE)
filels <- as.list(unzip(filetmp, list = TRUE))
filein <- as.character()
filesrc <- as.character()
for (i in 1:length(filels$Name)) 
    {
        filesrc <- paste(datadir, filels$Name[[i]], sep = "/")
        filein <- strsplit(filels$Name[[i]], ".", fixed = TRUE)[[1]][1]
        assign(filein, read.csv(filesrc))
    }
unlink(filetmp)
rm(filetmp, fileurl, filein, filesrc, i)
```
```{r}
str(activity)
```

* **activity** variable data type.
  + **steps**: integer class.
  + **date**: character class.
  + **interval**: integer class.
  
Take note that the variable **date** shall belonged to the date class data type instead of the current character class data type in  which date values are stored as character string. Time-series related analysis requires that the **date** variable be transformed into formal date class vector to warrant proper analysis.
&nbsp;

*5-numbers summary*
```{r}
summary(activity)
```
&nbsp;
*Missing value distribution*
```{r}
colMeans(is.na(activity))
```
<!-- workout the sum and mean of missing values for each column variable -->
```{r, echo = FALSE, results = "hide"}
na.count <- sum(is.na(activity$steps)) + 
            sum(is.na(activity$date)) + 
            sum(is.na(activity$interval))
na.ratio <- mean(is.na(activity$steps)) +
            mean(is.na(activity$date)) +
            mean(is.na(activity$interval))
print(na.count)
print(percent(na.ratio))


```

The 5-numbers summary of **activity** data set shows there are `r na.count` missing values (coded as NA), representing approx. `r percent(na.ratio)` missing values present in the data set. The ratio of missing values considered significant in this case which can skewed the outcomes of downstream analysis.

Missing value distribution, not in the usual sense of statistical distribution, simply reveals how the missing values are scattered within the data set.  
- **steps**: contains `r percent(colMeans(is.na(activity))[1])` missing value.  
- **date**: `r percent(colMeans(is.na(activity))[2])`.  
- **interval**: `r percent(colMeans(is.na(activity))[3])`.  

<!-- pre-processing data set  -->
```{r, message=FALSE, results="hide"}
# Pre-processing data for downstream analysis
names(activity) <- strCap(names(activity))  # capitalize variable names.
# hr <- activity$Interval %/% 100 # e.g. 2355 gives 23
# mn <- activity$Interval %% 100 # e.g 2355 gives 55
# transform Date variable to 'date' class and add Time variable of POSIXt
time.hms <- substr(as.POSIXct(sprintf("%04.0f", activity$Interval), format = '%H%M'), 12, 19)
date.time <- as.POSIXct(paste(activity$Date, time.hms))
activity <- activity %>%
            mutate(Date = as.Date(Date), Time = date.time) %>%
            arrange(Time)
# create daily activity summary
actsum <- activity %>% 
                select(Date, Steps) %>%
                mutate(Mean = as.numeric(0),
                       Median = as.numeric(0)) %>%
              group_by(Date) %>% 
             summarise(Mean = mean(Steps, na.rm = TRUE),
                       Median = median(Steps, na.rm = TRUE),
                       Steps = sum(Steps, na.rm = TRUE))
rm(time.hms, date.time)
```
&nbsp;

```{r}
h <- hist(actsum$Date, breaks = "days", freq = FALSE, plot = FALSE)
h$counts <- actsum$Steps
par(cex.axis = 0.8, las = 2)
plot(h, xaxt = "n", ann = FALSE, col = "grey")
axis(1, at = seq(h$breaks[1], h$breaks[length(h$breaks)], by = 5), 
     labels = c(format(as.Date(seq(h$breaks[1], h$breaks[length(h$breaks)], 
                                   by = 5 ), origin = "1970-01-01"), "%b-%d")))
abline(h = mean(actsum$Steps), col = "red")
abline(h = median(actsum$Steps), col = "green")
legend("topleft", legend = c("Mean", "Median"), 
       lty = c(1,1,1), col = c("red","green"), cex = 0.75)
title(main = "Number of Steps Taken Per Day", 
      sub = "( Year 2012 )",
      xlab = "Date", 
      ylab = "Daily Step Count",
      cex = 1)
```
<!-- histogram for Total Number of Steps Taken Per Day using ggplot2 graphic -->
<!-- NOT USE & HIDDEN FROM FINAL OUTPUT -->
```{r, message=FALSE, include=FALSE, echo=FALSE, results="hide"}
library(ggplot2, warn.conflicts = TRUE, quietly = TRUE)
# gr0 <- ggplot(actsum, aes(x = Date, y = Steps, fill = Steps))
gr0 <- ggplot(actsum, aes(x = Date, y = Steps))
gr0 <- gr0 + geom_col()
gr0 <- gr0 + scale_x_date(breaks = breaks_pretty(11))
gr0 <- gr0 + theme(plot.title = element_text(hjust = 0.5),
                   plot.subtitle = element_text(hjust = 0.5),
                   axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
gr0 <- gr0 + labs(title = "Number of Steps Taken Per Day",
                  subtitle = "(October & November, 2012)")
gr0 <- gr0 + xlab("Date") + ylab("Daily Step Count")
gr0 <- gr0 + geom_hline(data = actsum,
                        aes(yintercept = mean(Steps), col = "Mean"))
gr0 <- gr0 + geom_hline(data = actsum,
                        aes(yintercept = median(Steps), col = "Median"))
gr0 <- gr0 + scale_color_discrete(name = "Central Tendency")
print(gr0)
```

The total number of steps taken per day by the anonymous are illustrated in the histogram, where each bar in the graph represents the total step count recorded each day. The mean and median of the step count observations accumulated over the two months period as portrayed by the visual representation of the red and green horizontal lines overlaid on the histogram, hovering around the 10,000 mark indicate the anonymous individual performed, on average `r format(mean(actsum$Steps),nsmall=2)` steps, close to 10,000 steps per day in daily activities. The median value `r format(median(actsum$Steps),nsmall=2)` hovering slightly above the 10,000 mark. However, the daily mean and daily median are about 1000 steps apart from each other or about 10% of the mean/median indicate there is some abnormality in the data set.  
&nbsp;

<!-- 
frequency param setting for ts()
interval	frequency
annually        1
quarterly       4
monthly        12
weekly         52
daily         365.25
the frequency is the number of observations per cycle (normally a year, a week, a day or an hour).
if the frequency of observation is greater than once in a day, then there is more than one way of handling the frequency. e.g. data with per minute observation might have an hourly seasonality (frequency = 1 x 60), a daily seasonality (frequency = 1 x 60 x 24), a weekly seasonality (frequency = 1 x 60 x 24 x 7) and an annual seasonality (frequency = 1 x 60 x 24 x 365.25).
i.e.    frequency determines the number of observations per cycle, and therefore the number of cycle (number of observations / frequency) in the time series.
start param setting for ts()
i.e. start(m,n) where m = the number of the first cycle 
                      n = the starting nth incident for that first cycle
end(m,n) where m = the number of the last cycle
               n = the ending nth incident for that last cycle
      
e.g.
ts(actsum, 
        start = c(year(min(actsum$Date)), yday(min(actsum$Date))),
        end = c(year(max(actsum$Date)), yday(max(actsum$Date))),
        frequency = 365)
ts(activity,
        start = c(1,1),
        end = c(17568/(12*24), 12*24),
        frequency = 12*24)
-->

### What is the Average Daily Activity Pattern?

<!-- Average Daily Activity Pattern plot using base graphic -->
<!-- Not Use and Hidden from  final output -->
```{r, include=FALSE, echo=FALSE, results="hide"}
pal <- c(rgb(0,0,1), rgb(0,1,0), rgb(1,0,0), rgb(1,0,1), rgb(1,1,0), rgb(0,1,1))
activity.ts <- aggregate(Steps ~ Interval, data = activity, FUN = mean)
par(cex.axis = 0.8, las = 2)
plot(activity.ts, type = "l", col = "royalblue3")
h <- activity.ts[which(activity.ts$Steps == max(activity.ts$Steps)),]
abline(v = h$Interval, col = pal[3])
title(main = "Average Daily Activity Pattern", 
      # sub = "( data points plotted @ 5-min interval )",
      xlab = "Time Interval", 
      ylab = "Step Count",
      cex = 0.9,
      cex.sub = 0.7)
```
<!-- Time-series plot for Average Daily Activity Pattern using base graphic  -->
```{r}
pal <- c(rgb(0,0,1), rgb(0,1,0), rgb(1,0,0), rgb(1,0,1), rgb(1,1,0), rgb(0,1,1))
activity.ts <- ts(aggregate(Steps ~ Interval, data = activity, FUN = mean))
par(cex.axis = 0.8, las = 2, ann = FALSE)
plot(activity.ts[,2] ~ activity.ts[,1], type = "l", col = "royalblue3")
h <- activity.ts[which(activity.ts[,2] == max(activity.ts[,2])),]
abline(v = h[1], col = pal[3])
title(main = "Average Daily Activity Pattern", 
      xlab = "Time Interval", 
      ylab = "Average Step Count",
      cex = 0.9,
      cex.sub = 0.7)
```

The average daily activity pattern time-series plot shows the anonymous individual's daily activity pick up pace starting 0600 hour lasted for 13 hours, paces begin to wind-down from 1900 hour onward and gradually slow down to negligible level at around 2130 hour continuing forward. There is a noticeable sharp spike (red color vertical line) in the average activity at `r sprintf("%04.0f", h[1])` hour to `r format(h[2],nsmall=2)` steps, which is double the maximum average step count of 100 for most part of the day.
&nbsp;

```{r}
par(mfrow = c(1,2))
boxplot(activity.ts[,2], ylab = "average step count by interval")
hist(activity.ts[,2], main = NULL, xlab = "average step count by interval")
```

As shown in the Boxplot (left) and histogram (right) above, least number of data point or lowest frequency beyond the average 200-step count mark affirmed the unusual sharp spike exhibited in the Average Daily Activity Pattern time series graph plot.  
&nbsp;

### Imputing Missing Values

As mentioned earlier, there are `r na.count` or `r percent(na.ratio)` missing values (coded as NA) found in the **activity** data set. 

- **steps** variable: contains `r percent(colMeans(is.na(activity))[1])` missing value.  
- **date** variable: `r percent(colMeans(is.na(activity))[2])`.  
- **interval** variable: `r percent(colMeans(is.na(activity))[3])`.  

The present of missing values will induce biased analysis outcomes. Therefore K-Nearest Neighbour (knn) algorithm is employed to impute missing values in the data set, as knn imputation method is well accepted in the community among the practitioners.

<!-- 
Generally accepted approach to determine k-value for knn.impute() base on internet search.
k = sqrt(17568)/2 
where number of observations = 17568
-->
<!-- Imputing missing values in data set using knn.impute() -->
```{r, message=FALSE, results = "hide"}
# knn.impute() function
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("impute")
library(impute)
activity.imputed <- impute.knn(data.matrix(activity[,1:3]), k = 65)$data
activity.imputed <- as.data.frame(activity.imputed)
activity.imputed <- mutate(activity.imputed, Time = activity$Time)
# prepare daily summary of imputed activity data set
actsum <- activity.imputed %>% 
                select(Date, Steps) %>%
                mutate(Date = as_date(Date),
                       Mean = as.numeric(0),
                       Median = as.numeric(0)) %>%
              group_by(Date) %>% 
             summarise(Mean = mean(Steps),
                       Median = median(Steps),
                       Steps = sum(Steps))
```